Conference Paper Title*\\ {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and should not be used} \thanks{Identify applicable funding agency here. If none, delete this.}

### 2023-05-20 LARS_paper/paper.tex
Fix spellings

\section{O Robô Caramelo}

According to the concepts presented, a small quadruped was designed and developed for research and educational purposes.(figuras \ref{fig:terrenos}, \ref{fig:caramel_body} e \ref{fig:moving_body}).
During the prototyping phase, the hardware architecture and physics structure was designed. Following that, the structure of the robot was printed and integrated with the hardware components. The robot's actuators are servo motors of the model dynamixel MX-28, and the central processing is composed of a Raspberry PI 4. Furthermore, it contains an inertial sensor, model MPU6050. All the software was developed with the Robot Operating System 2 Humble (ROS2) \cite{ROS2Humble}.

The robot's structure is mammalian-type, characterized by the presence of two joints in the sagittal plane, similar to animals such as dogs and horses.  The robot’s legs follow a configuration called "full-elbow", in which the four legs are oriented backward, such as the elbows. Furthermore, it features 3 DoFs per leg, providing the legs substantial freedom of movement. The robot was designed to favor the mass balance between the body and the robot's legs, that is, the most of mass is in the body or closer to it. Lighter legs can move quickly without change, significantly, the center of gravity of the robot, which increases the stability and requires less control complexity.  On the other hand, the legs should be strong enough to support the robot's weight and the impact of the legs on the ground.

The internal components, which include sensors, processing units, and the communication interface with the actuators, were arranged symmetrically in the robot's torso. The actuators (components that contribute to the most of system mass) were positioned as close as possible to the body. A highlight for the motor that acts on the tibia joint was its installation in the upper part of the femur, in order to reduce the moment of inertia of the leg. This choice required the addition of a transmission system between the engine and the tibia link, consisting of a rigid metal rod with a spheric joint at each edge.

The gait developed to the robot follows the sequence of phases demonstrated in figure \ref{fig:trot_pattern}, in which the blank rectangles represent the  swing phase and the gray rectangles represent the stance phase. During the stance phase, the legs are on the ground and push the robot in the desired direction.  In the swing phase, the legs are raised to move the legs to the next point of support. This is the main gait of the robot, called a trot, based on the concept of periodic and symmetrical gaits. The periodic gaits are characterized by the continuous repetition of the same movements in the same instants. While the symmetry is a characteristic of gaits that move a pair of legs together, leaving and returning to the ground in a synchronized way. Although the structure of the robot makes it possible to perform other types of gaits, in this work, only the trot was considered. Still,  in order to reduce the complexity of locomotion control, a discontinuous gait was adopted, that is, the robot's body moves only when all the legs are on the ground, performing an intermittent movement. In figure \ref{fig:trot_pattern}, it is possible to realize that always the same pair of diagonal legs move at the same instant. Between two consecutive phases of swing, there is a moment when all legs are in stance, which is when the robot body is displaced in the desired direction of locomotion.

@@

The control subsystem of the robot is composed of two main components: the rotation control and the gait planner. The rotation control of the body is responsible for controlling the pitch and angles of the robot, in order to keep it at 0 degrees at all times. The gait planner is responsible for controlling each leg and, consequently, the robot's body. It defines where the foot will step and where the body should move based on the swing and stance steps. The gait planner uses a trajectory planner to calculate the trajectory the foot should take in space in order to move the robot in the desired direction and speed. The points of this trajectory are converted into the angles of the joints through the kinematic model. The kinematic model of the robot is responsible for mapping the three-dimensional position of each leg and obtaining the joints' angles that compose it.  In addition, the trajectory planner also considers the control effort sent by the rotation controller.

The gait planner sends the points at which each foot must be at a frequency of $\SI{50}{\hertz}$. This frequency was defined as being the highest that the robot processing system was able to support and by respecting the Nyquist theorem, which indicates that in this case, the frequency should be greater than $\SI{4}{\hertz}$ -- since the step period is $0.5 s$.

@@

The trajectory planner is responsible for calculating the trajectory that each foot must perform, both in the \textit{stance} and \textit{swing} phases. For Caramel, the trajectory is a cycloidal curve in both stages. As shown in \cite{Shi2021}, a cycloidal curve can be defined in the three-dimensional space between the points $(x_o, y_o, z_o)$ and $(x_f, y_f, z_f)$ as a function of time, $t$, by equations (\ref{eq:traj_x}) to (\ref{eq:traj_k}), with $H$ the step height and $T$ the period.
\begin{equation}
  x = (x_f - x_o) \frac{K - \sin{(K)}}{2 \pi} + x_o
  \label{eq:traj_x}
\end{equation}
\begin{equation}
  y = (y_f - y_o) \frac{K - \sin{(K)}}{2 \pi} + y_o
  \label{eq:traj_y}
\end{equation}
\begin{equation}
  z = H \frac{1 - \cos{(K)}}{2} + z_o
  \label{eq:traj_z}
\end{equation}
\begin{equation}
  K = \frac{2 \pi t}{T}
  \label{eq:traj_k}
\end{equation}

The graph of a cycloidal curve in 3D space is shown in figure \ref{fig:traj_space}. The same trajectory is illustrated in figure \ref{fig:traj_time} as a function of time. It is possible to notice that the curve has the first null derivative at the time when the foot touches the ground, which is favorable to open loop control, since the softer the landing of the foot, the fewer disturbances are caused in the system.

\begin{figure*}[h]
  \centering
  \begin{subfigure}[t]{0.32\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{Cycloid_space.png}
    \caption{Curva no espaço 3D.}
    \label{fig:traj_space}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{Cycloid_time.png}
    \caption{Curva no tempo.}
    \label{fig:traj_time}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{Cycloid_modified.png}
    \caption{Curva com disposição de pontos modificada.}
    \label{fig:traj_time_modified}
  \end{subfigure}
  \vfill
  \caption{Trajetórias cicloidais para o passo de robô.}
  Fonte: autores.
  \label{fig:traj_curve}
\end{figure*}
In addition to height, the distances in $x$ e $y$, and period, the robot's trajectory planner also counts upon two parameters in order to further improve the control of the force with which the foot touches the ground. Since the robot’s actuators are position-controlled servo motors, the torque is proportional to the displacement it must perform between the points of the trajectory. That is, the higher the resolution of the trajectory, the smoother the movement. However, the resolution of the $N$ trajectory is fixed, given as a function of the step period and the frequency control of the gait planner ($\SI{50}{\hertz}$) $N = 50T$. Therefore, the strategy adopted is to space the same amount of points unevenly throughout the step period, so that there are more points near the moment the foot lands on the ground and less points near the moment the foot is erected. The $P_T$ parameter is a fraction of the total step period, and the $P_N$ is the fraction of the total number of points that should be between the $0$ and $P_T cdot T$. In other words, if $P_T = 0,66$ e $P_N = 0,33$, $33\%$ of $N$ will be in the first two-thirds of the period, while the remaining $67\%$ will be in the final one-third. The trajectory, considering these parameters, is illustrated in figure \ref{fig:traj_time_modified}.

>>>>

\section{Robot Development}

A small quadruped was designed and developed for research (Fig. X).
During the prototyping phase, the hardware architecture and physics structure was designed. Following that, the structure of the robot was printed and integrated with the hardware components. The robot's actuators are servo motors of the model dynamixel MX-28, and a Raspberry Pi 4 was chosen as processing unit. Furthermore, it contains an inertial sensor, model MPU6050. All the software was developed with the Robot Operating System 2 Humble (ROS 2) \cite{ROS2Humble}.

IMAGEM DO ROBO

The robot's structure is mammal-type, which is characterized by the presence of two joints in the sagittal plane, similar to animals such as dogs and horses.  The robot’s legs follow a configuration called full-elbow, in which the four legs are oriented backward, as shown in Fig. X. Furthermore, it features 3 degrees of freedom (DoF) per leg, providing the legs with substantial freedom of movement. The robot was designed to favor the mass balance between the body and the robot's legs, that is, the majority of its mass is in located in the body or as close to it as possible. Lighter legs can move quickly without significantly changing the center of gravity of the robot, which increases the stability and requires less control complexity.  On the other hand, the legs should be strong enough to support the robot's weight and the impact with the ground.

The internal components, which include sensors, processing unit, and the communication interface with the actuators were arranged symmetrically in the robot's body. The actuators (components that contribute with most of the system's mass) were positioned as close as possible to the body. The motor that acts on the tibia joint was placed in the upper part of the femur, in order to reduce the moment of inertia of the leg. This choice required the addition of a transmission system between the engine and the tibia link, consisting of a rigid metal rod with a ball joint at each edge.

Quadruped robots move according to a gait pattern. The gait implemented for the robot follows the sequence of phases demonstrated in Fig. \ref{fig:trot_pattern}, in which the white rectangles represent the  swing phase and the gray rectangles represent the stance phase. During the stance phase, the legs are on the ground and push the robot in the desired direction.  In the swing phase, the legs are raised and moved to the next support point. This gait pattern is called a trot, and is based on the concept of periodic and symmetrical gaits. The periodic gaits are characterized by the continuous repetition of the movements in the same instants. The symmetry is a characteristic of gaits that move a pair of legs together, leaving and returning to the ground in a synchronized way. Although the structure of the robot makes it possible to perform other types of gaits, in this work, only the trot was considered. In order to reduce the control complexity, a discontinuous gait was adopted, that is, the robot's body moves only when all the legs are on the ground, resulting in an intermittent movement. In Fig. \ref{fig:trot_pattern}, it is noticeable that the same pair of diagonal legs always move at the same instant. Between two consecutive swing phases, there is a moment when all legs are in stance, which is when the robot body is displaced in the desired direction of locomotion.

@@

The control subsystem of the robot is composed of two main components: the body's stabilization controller and 
joints' controllers. The body's stabilization controller is responsible for controlling the pitch and roll angles of the robot, maintaining it at 0 degrees. Meanwhile, the joints' controllers, each one a PID, are responsible for controlling each joint of the robot independently.

The gait planner is responsible for controlling each leg and the robot's body. It defines where each foot will step and where the body should move to based on the desired velocity command. The gait planner uses a trajectory planner to calculate the trajectory each foot and body should follow. This trajectory is calculated based on the velocity command. The points of this trajectory are converted into the angles of the joints through the kinematic model. The kinematic model of the robot is responsible for mapping the three-dimensional position of the foot to the joints' angles of the respective leg.

The gait planner sends the points at which each foot must be at a frequency of $\SI{50}{\hertz}$. This frequency was defined as being the highest that the robot processing system was able to support and by respecting the Nyquist theorem, which indicates that in this case, the frequency should be greater than $\SI{4}{\hertz}$ --- since the step period is $0.5 s$.

% TODO ref nyquist

@@

The trajectory planner is responsible for calculating the trajectory that each foot must perform. For both stance and swing phases, a cycloidal curve is adopted. As shown in \cite{Shi2021}, a cycloidal curve can be defined in the three-dimensional space by (\ref{eq:traj_x}) to (\ref{eq:traj_k})
between the points $(x_o, y_o, z_o)$ and $(x_f, y_f, z_f)$ as a function of time $t$, step height $H$, and period $T$.
\begin{equation}
  x = (x_f - x_o) \frac{K - \sin{(K)}}{2 \pi} + x_o
  \label{eq:traj_x}
\end{equation}
\begin{equation}
  y = (y_f - y_o) \frac{K - \sin{(K)}}{2 \pi} + y_o
  \label{eq:traj_y}
\end{equation}
\begin{equation}
  z = H \frac{1 - \cos{(K)}}{2} + z_o
  \label{eq:traj_z}
\end{equation}
\begin{equation}
  K = \frac{2 \pi t}{T}
  \label{eq:traj_k}
\end{equation}

The graph of a cycloidal curve in 3D space is shown in Fig. \ref{fig:traj_space}. The same trajectory is illustrated in Fig. \ref{fig:traj_time} as a function of time. It is noticeable that the curve has the first null derivative when the foot touches the ground, which is favorable to open loop control, since the smoother the landing, the fewer disturbances are caused in the system.

\begin{figure*}[h]
  \centering
  \begin{subfigure}[t]{0.32\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{Cycloid_space.png}
    \caption{Curva no espaço 3D.}
    \label{fig:traj_space}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{Cycloid_time.png}
    \caption{Curva no tempo.}
    \label{fig:traj_time}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{Cycloid_modified.png}
    \caption{Curva com disposição de pontos modificada.}
    \label{fig:traj_time_modified}
  \end{subfigure}
  \vfill
  \caption{Trajetórias cicloidais para o passo de robô.}
  Fonte: autores.
  \label{fig:traj_curve}
\end{figure*}

In addition to the step height, the $x$ and $y$ translations, and the period, the robot's trajectory planner also considers two parameters to make the landing of the foot smoother. Since the robot’s actuators are servo motors controlled by position, the torque is proportional to the displacement it must perform between the points of the trajectory. Therefore, the higher the resolution of the trajectory, the smoother the movement. However, the resolution of the trajectory $N$ is fixed, given as a function of the step period and the frequency of the gait planner ($\SI{50}{\hertz}$) $N = 50T$. Thus, the strategy adopted is to distribute the $N$ points unevenly throughout the step period, so that there are more points near the moment the foot lands on the ground and less points near the moment the foot takes off. The $P_T$ parameter is a fraction of the total step period, and the $P_N$ is the fraction of the total number of points that should be between $0$ and $P_T \cdot T$. In other words, if $P_T = 0.66$ and $P_N = 0.33$, $33\%$ of $N$ will be in the first two-thirds of the period, while the remaining $67\%$ will be in the final one-third. The trajectory, considering these parameters, is illustrated in Fig. \ref{fig:traj_time_modified}.
